---
title: "R nf-core utils tutorial"
author: "Louis Le Nezet"
date: "30/03/2026"
url: "https://github.com/nf-core/r-nf-core-utils"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
        fig_crop: no
header-includes: \usepackage{tabularx}
vignette: |
    %\VignetteIndexEntry{R nf-core utils tutorial}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
        wrap: 80
---

```{r width_control, echo = FALSE}
options(width = 100)
```

# Introduction

This package is meant to be use inside Nextflow `template()` script.
Its aim is to provide useful function to take care of the connection between
Nextflow variable and R logic.

## Main function

There is two important function in this package.

### Function `process_inputs()`

This function takes as inputs a list of expected options with their default values,
the argument string and different validation rules.

#### Parameter `opt`

This parameter should list all the different variable that you might use in the
main script. You can set them to a default value or event directly initialize them
with Nextflow variable such as:

```{r, eval = FALSE}
opt <- list(
    prefix = "${task.ext.prefix}",
    seed = 1
)
```

#### Parameter `args`

The argument string correspond to `${task.ext.args}` in Nextflow and will be parsed
with `parse_arguments()`. This function expect all arguments to be in the form
`--key value`.

If you need spaces for one value, use bracket around it, such as `--key "value with space"`.
All the key / value pairs will then overwrite their counter part in the options list
passe to `process_inputs()`.

#### Validation rules

The `process_inputs()` function will enforce the following rules to the keys listed:

- `keys_to_nullify`: will be set to R `NULL` value if is "null" or empty
- `expected_files`: these paths should be existing files
- `expected_folders`: these path should be existing folder
- `expected_double`: these values will be converted with `as.double()` or should be `NULL`
- `expected_integer`: these values will be converted with `as.integer()` or should be `NULL`
- `required_opts`: these keys should be non-null values

### Function `process_end()`

This function will emit a `versions.yml` and a `R_sessionInfo.log` file in the directory
provided. The version file will be populated with the R version, the version of nfcore.utils
and the version of the additional packages given.

#### Parameter `packages`

This parameter should be a named list where the name correspond to the conda package name
and the value the R package name.

Such as:

```{r, eval = FALSE}
process_end(
    packages = list(
        "r-stats" = "stats"
    ),
    task_name = "${task.process}"
)
```

## Usage example

If we take for example the [`custom/geneticmapconvert` process in nf-core modules](https://nf-co.re/modules/custom_geneticmapconvert/).

The nextflow process is the following:

```{groovy, eval = FALSE}
process CUSTOM_GENETICMAPCONVERT {
    tag "$meta.id"
    label 'process_single'

    input:
    tuple val(meta), path(map_file)

    output:
    tuple val(meta), path("${prefix}.glimpse.map"), emit: glimpse_map
    path "versions.yml", emit: versions_geneticmapconvert, topic: versions

    when:
    task.ext.when == null || task.ext.when

    script:
    prefix = task.ext.prefix ?: "${meta.id}"
    args = task.ext.args ?: ''

    """
    echo ${args} // In the form --tolerance 0.15
    """

    template 'geneticmapconvert.R'
}
```

Then in the `templates/geneticmapconvert.R` we use the following

```{r, eval = FALSE}
library(nfcore.utils)
library(data.table)
library(stringr)

### INPUTS PARSING ###
opt <- list(
    map_file = "${map_file}",
    chr = "${meta.chr}",
    prefix = "${prefix}",
    tolerance = NULL
)

process_input(
    opt = opt,
    args = "${args}",
    keys_to_nullify = c("prefix", "tolerance"),
    expected_files = c("map_file"),
    expected_double = c("tolerance"),
    required_opts = c("map_file", "prefix")
)

### MAIN SCRIPT ###

...

### END of PROCESS ###
process_end(
    packages = list(
        "r-data.table" = "data.table",
        "r-stringr" = "stringr"
    ),
    task_name = "${task.process}"
)
```

# Session information

```{r}
sessionInfo()
```
